---
title: "Class 12: Q13-Q14 Homework"
author: 'Pamelina Lo (SID: A16735368)'
date: "2024-11-07"
output: pdf_document
---

# Section 1. Proportion of G/G in a population

Downloaded a CSV file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39780097-40010098;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

Lets look at a different population. I picked the GBR (Great Britian). 

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)
```

Find proportion of G|G

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

This variant that is assoicated with childhood asthma is more frequent in the GBR population than the MKL population.



# Section 4: Population Scale Analysis

**Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.** 

```{r}
expr <- read.table("https://bioboot.github.io/bimm143_F24/class-material/rs8067378_ENSG00000172057.6.txt")
head(expr)
```

Sample Size:
```{r}
nrow(expr)
```

**Sample size and Median expression levels for each genotype:**
```{r}
library(dplyr)

genotype_summary <- expr %>%
  group_by(expr$geno) %>%
  summarise(
    sample_size = n(),
    median_expression = median(exp, na.rm = FALSE)
  )

print(genotype_summary)
```

The sample sizes: A|A = 108 ,  A|G = 233,  G|G = 121
Median Expression: A|A = 31.25 , A|G = 25.06 , G|G = 20.07


**Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?** 

I can infer that the relative expression value is that there is more expression of the A|A genotype, than G|G because, based on the plots, A|A appears to be higher than G|G plot because its median is larger and the distribution of the plot is more spread out. The SNP does effect the expression of ORMDL3 because having G|G genotype in this location associates to a reduced expression on this gene.  

```{r}
library(ggplot2)
```

```{r}
ggplot(expr) +
  aes(geno, exp, fill = geno) +
  geom_boxplot()
```

